The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2022-01-14 paper/paper.Rmd
Add the missing code chunk.
revised introduction

---
title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: ""
    location: ""
    email: ""
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
  - \usepackage{color, colortbl}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;"
abstract: > 
 The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets and the code to produce them are provided in an accompanying open source R package, called [CENSORED]. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users. 
---

@@

"Open data" is data that is freely accessible, modifiable, and shareable by anyone for any purpose [@opendata]. This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call "wild data" because it requires substantial cleaning and tidying to tame it into textbook shape. @HuebnerMariannePhD2016Asat emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building <!-- AE comment: It's unclear from context what is being referred to from the Huebner et al 2020 citation -->  [@HuebnerMarianne2020Haar]. 
Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo]. In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. @DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected [@Chatfield1985TIEo]. There are few research papers that document the data cleaning [@WickhamHadley2014TD]. Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties [@HuebnerMarianne2020Haar].

This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN]. In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy]. According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].

@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. <!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar. It sounds like this dataset is used in many books, why choose that one? The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was. From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset. Be explicit about this.-->
Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda. We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race [@racismnotrace].  <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->

@@

>>>>

---
title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: Telethon Kids Institute
    location: Nedlands, WA 6009
    email: ""
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
  - \usepackage{color, colortbl}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;"
abstract: > 
 The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets and the code to produce them are provided in an accompanying open source R package, called [CENSORED]. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users. 
---

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was developed into a case study for use in teaching exploratory data analysis at Iowa State University. The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.

This disparity between the average and the individual is a part of statistics that is given less attention. It is this disparity which makes this particular data set very important as a textbook example. Textbook data sets have staying power if it has a challenge that has never been completely solved. The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?

For the field of statistics, and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy or statistics for the public. The two are not the same. As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, the message for the individual is that you are more than your demographic. The majority of people tracked are not like the average, the average is an anomaly. If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.

As a textbook data though, the dates are tired. The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students. Using tired old textbook data sets sets up a misconception that the field is not current.  The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This is our story of the journey from open [@opendata] but wild data to textbook data, it's not entirely pretty.

@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN]. In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy]. According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].  <!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar. It sounds like this dataset is used in many books, why choose that one? The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was. From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset. Be explicit about this.-->
Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda. We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the *purposes of studying racism, not race* [@racismnotrace].  <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->

<!-- 
"Open data" is data that is freely accessible, modifiable, and shareable by anyone for any purpose [@opendata]. This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call "wild data" because it requires substantial cleaning and tidying to tame it into textbook shape. @HuebnerMariannePhD2016Asat emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building --> <!-- AE comment: It's unclear from context what is being referred to from the Huebner et al 2020 citation  [@HuebnerMarianne2020Haar]. --> 
<!-- Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo]. In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. @DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task. -->

<!-- Despite it's importance, this IDA stage is often undervalued and neglected [@Chatfield1985TIEo]. There are few research papers that document the data cleaning [@WickhamHadley2014TD]. Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties [@HuebnerMarianne2020Haar]. -->

This paper demonstrates the steps of cleaning data, and documents the process, as recommended by @HuebnerMariannePhD2016Asat.  They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo]. In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. @DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.

@@

```{r tidy-rate-hour}
```
